Acceleration of Bilateral filtering algorithm for manycore and multicore architectures

### 2012-03-11 trunk/LateX/paper8.tex
Added and changed few plots, rearranged a few sections.

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering infinite-resolution images, and bilateral filtering in the digital domain requires a finite set of pixels as the input image. Therefore, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e., one calculation per Neighbor and one set of calculations per Center Pixel. A formal expression is given below:

\begin{equation}
\label{eq:depen}
filtered\: pixel\: x=\frac{\substack{N\\
\sum\\
i=1
}
f(\xi_{i})\cdot c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}{\substack{N\\
\sum\\
i=1
}
c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}
\end{equation}

In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter. Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated. Consequently, the Bilateral filter has two data dependencies. As explained later in this paper, our pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.

Another observation taken into account is that the values of Equation~\ref{spatial}, for the Neighborhood, can be computed in advance as the values depend on a simple Gaussian probability distribution function. However, the values of Equation~\ref{photometric} and consequently of Equation~\ref{combined} cannot be known a priori as they depend on the similarity of signals between the current Center Pixel and the Neighborhood. Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window. By reason of the foregoing, the bulk of the time spent in processing an image using the Bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.

\subsection{A Naive Approach to Bilateral filtering kernel}

\begin{figure}
\centering\includegraphics[scale=0.35]{images/the1}
\includegraphics[clip,scale=0.35]{images/the202}\includegraphics[scale=0.35]{images/the2}\includegraphics[scale=0.35]{images/the3}
\caption{Example of a naive Bilateral filter iterations}
\end{figure}

The naive stencil algorithm is the standard model that is predominantly used in Bilateral filtering kernel. Previous implementations\cite{Paris2009,Pham2005} of Bilateral filtering kernel improves the hardware dependent aspects and hence expend resources on achieving performance gains that can not match the effort taken. Here, we address a root problem of the standard naive algorithm by proposing a new algorithm, and then develop hardware dependent improvements for further speedup. Algorithm~\ref{alg:Example-Algorithm-float} is the pseudo-code for the standard naive Bilateral filtering kernel.

%\begin{algorithm}[!h]
%\caption{Generalized Naive Stencil Kernel Algorithm}
%\label{alg:Example-Algorithm-float}
%\begin{algorithmic}
%\FORALL{element  $s_{i,j}$ in source image s}
%\FORALL{data dependency $n$}
%\STATE{$X_n \leftarrow 0$}
%\ENDFOR
%\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
%\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
%\FORALL{data dependency $n$}
%\STATE{$X_n \leftarrow X_n + f_n(s[i,j],s[i+k,j+l])$}
%\ENDFOR
%\ENDFOR
%\ENDFOR
%\STATE{output[$i,j] \leftarrow f_{output}(X_{1}, \dots, X_{last})$}
%\ENDFOR
%\end{algorithmic}
%\end{algorithm}

\begin{algorithm}[!h]
\caption{Naive Bilateral Filter Algorithm}
\label{alg:Example-Algorithm-float}
\begin{algorithmic}
\FORALL{element  $s_{i,j}$ in source image s}
\STATE{$X_{1} \leftarrow 0$}
\STATE{$X_{2} \leftarrow 0$}
\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\ENDFOR
\ENDFOR
\STATE{output[$i,j] \leftarrow X_{1} / X_{2}$}
\ENDFOR
\end{algorithmic}
\end{algorithm}

\subsection{Pair-symmetric Bilateral filtering kernel}
The pair-symmetric algorithm, shown in Algorithm~\ref{alg:Example-Algorithm-float-1}, reduces the amount of Center-Neighbor computations by half. Thus, it can theoretically cut the processing time in half at the cost of increased memory requirement, which increases with the amount of dependencies. Its implementation is non-trivial due to potential race-conditions.

Our pair-symmetric Bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel, and that the codes are symmetric with respect to Center-Neighbor paired computations. For example, in the Bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$. It then follows that the naive Bilateral filtering algorithm dictated in Algorithm~\ref{alg:Example-Algorithm-float} above, although simple, performs twice as many calculations than necessary.

The Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive Bilateral filtering algorithm. Instead of reading from all Neighbors, our pair-symmetric algorithm reads from only half of the Neighbors. The result of each Center-Neighbor paired computation is then added to the $X_n$ register, which eventually obtains the result of data dependency $n$. This is trivial as the naive Bilateral filtering algorithm already does this. What our pair-symmetric algorithm does unique is that the result of each Center-Neighbor paired computation is also added to and stored in the Neighbor's position in a temporary array ("Map'') that is used for storing the results of a data dependency. So, there are $n$ Maps for $n$ number of dependencies.

After the data dependency conditions are met, the function that returns the output data from the dependencies iterates over the Maps. As applied to the Bilateral filter, the data dependency lies in calculating each pixel's numerator and denominator that is used to derive the pixel's output result. After the numerators and denominators are derived, a function iterates and divides the corresponding elements of the two arrays to return the output.

%\begin{algorithm}[h]
%\caption{Pair-symmetric Stencil Algorithm}
%\label{alg:Example-Algorithm-float-1}
%\begin{algorithmic}
%\STATE /* D is the set of dependency arrays */
%\FORALL{data dependency $n$}
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$D_n[i,j] \leftarrow 0$}
%\ENDFOR
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image $s$}
%\FOR{$k=1$ \TO $filter\_hw$}
%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
%\FORALL{data dependency $n$}
%\STATE $temp \leftarrow f_n(s[i,j],s[i+k,j+l])$
%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
%\STATE $D_n[i+k,j+l] \leftarrow D_n[i+k,j+l] + temp$
%\ENDFOR
%\ENDFOR
%\ENDFOR
%\FOR{$l = -filter\_hw$ \TO $0$}
%\FORALL{data dependency $n$}
%\STATE $temp \leftarrow f_n(s[i,j],s[i,j+l])$
%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
%\STATE $D_n[i,j+l] \leftarrow D_n[i,j+l] + temp$
%\ENDFOR
%\ENDFOR
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$output[i,j] \leftarrow f_{output}(D_{1}[i,j], \dots, D_{last}[i,j])$}
%\ENDFOR
%\end{algorithmic}
%\end{algorithm}

%\begin{algorithm}[h]
%\caption{Pair-symmetric Bilateral Filtering kernel}
%\label{alg:Example-Algorithm-float-1}
%\begin{algorithmic}
%\STATE /* D is the set of dependency arrays */
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$D_1[i,j] \leftarrow 0$}
%\STATE{$D_2[i,j] \leftarrow 0$}
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image $s$}
%\FOR{$k=1$ \TO $filter\_hw$}
%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
%\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i+k,j+l]$}
%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
%\STATE{$D_1[i+k,j+l] \leftarrow D_1[i+k,j+l] + temp \cdot s[i,j]$}
%\STATE{$D_2[i+k,j+l] \leftarrow D_2[i+k,j+l] + temp$}
%\ENDFOR
%\ENDFOR
%\FOR{$l = -filter\_hw$ \TO $0$}
%\STATE{$temp \leftarrow e^{C_{1}l^2 + C_2(s[i,j] - s[i,j+l])^{2}}$}
%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i,j+l]$}
%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
%\STATE{$D_1[i,j+l] \leftarrow D_1[i,j+l] + temp \cdot s[i,j]$}
%\STATE{$D_2[i,j+l] \leftarrow D_2[i,j+l] + temp$}
%\ENDFOR
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$output[i,j] \leftarrow D_{1}[i,j] / D_{2}[i,j]$}
%\ENDFOR
%\end{algorithmic}
%\end{algorithm}

@@

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work should be arranged such that warps can work as independently from each other as possible in order to minimize explicit synchronization calls, as illustrated in Figure~\ref{fig:pspro}. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work of each warp on its current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call would result in warps trespassing into the territory of other warps, thereby causing race conditions as threads from different warps would be consequently prone to writing to the same memory location.

\subsection{Shared Memory and Tile Processing}
\begin{figure}
\includegraphics[width = 0.48 \columnwidth]{images/project24.pdf}
\caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
\label{fig:pspro}
\end{figure}

\begin{figure}
\includegraphics[width = 0.48  \columnwidth]{images/tiles19.pdf}
\caption{Illustration of our CUDA implemented pair-symmetric algorithm processing a 24 x 12 pixel image using 2 x 2 thread-blocks each with 4 x 3 threads and with the half-width parameter set to 3 pixels.}
\label{fig:pspro}
\end{figure}

\begin{figure}
\includegraphics[width = 0.48  \columnwidth]{images/smallfaces5.pdf}
\caption{An illustrated overview of the bilateral filtering process. The six raster images of processed tiles reflect the successive work of the four concurrent thread-blocks iteratively carried out over the padded source-image.}
\label{fig:pspro}
\end{figure}

CUDA features three types of memory with read/write capability suitable for general-purpose computing on GPU: \begin{inparaenum}[(i)] \item global memory, which is accessible by all threads in all thread blocks, \item shared memory, which is provided separately for each block of threads to share among themselves, and \item registers, which are separate and distinct for each thread and of which each is only accessible by the thread to which it belongs\end{inparaenum}. Of these three types of memory, the slowest is global memory, which unfortunately is where the host must first send the input data before the GPU can begin its work. So, in order to see any meaningful performance in CUDA, data stored in global memory must be transferred to faster memory--i.e., shared memory, which is hundreds of times faster. However, the trade-off is that the shared memory size is much smaller than the global memory size, and this fact compels a tile-based approach to the implementation of Bilateral filtering algorithm.

Our tile-based processing approach is self-explanatory. Equal sized partitions of the input data are processed, and then any remaining data on and between the edge and the last partition that is closest to (but not falling over) the edge boundary of the 2-dimensional data representation is processed. With respect to the naive Bilateral filtering algorithm, tile-based processing is trivial and needs no further explanation because simply partitioning the space and processing it will suffice. However, for the pair-symmetric algorithm that we propose, tile-based Bilateral filtering algorithm requires adjustments in the order in which tiles are processed in order to guarantee the exclusion of race conditions.

\begin{figure}
\includegraphics[width=0.48\textwidth]{images/stackedtiles}
\caption{Tiling for Pair-Symmetric Stencil Processing with each tile having a unique color and overlapping tiles have a resultant color that results from mixing colors of overlapping tiles. The pair-symmetric thread access pattern ensures exclusive memory writes.}
\label{fig:Tiling-for-Pair-Symmetric}
\end{figure}

\begin{figure}
\includegraphics[width=0.48\textwidth]{images/lenatiles}
\caption{Input image after one iteration of pair-symmetric algorithm. Tiles do not have perfect boundaries due to overlaps.}
\label{fig:An-iteration-of}
\end{figure}

If processes were assigned to adjacent tiles, race conditions would occur because the edge cells of the tiles would overlap each other due to intersecting half-width periphery regions, as illustrated in Figure~\ref{fig:Tiling-for-Pair-Symmetric}. To get around this problem, simultaneously processed tiles can be spaced apart in strides such that no two tiles being processed at the same time intersect with one another. An example of such a technique is illustrated in Figure~\ref{fig:An-iteration-of} (the tiles are flipped sideways in this rendering).

@@

\subsection{Memory to computation ratio of Bilateral filtering kernel}
\begin{figure}[h!]
\begin{centering}
\subfigure{ \includegraphics[width=0.48\textwidth]{images/noexplarge}} 
\par\end{centering}
\caption{Percentage of time taken by sequential kernel when run without $exp$ instruction, $filter$ $radius$=10.}
\label{fig:noexp} 
\end{figure}

Unlike traditional regular embarrassingly parallel stencil kernel codes where memory management plays a pivotal role in realizing performance, Bilateral filtering kernel is extremely computationally intensive with only a meager amount of time spent in memory accesses. Bilateral filtering kernel requires computation of $exponential$ function ($exp$ function) which is a major component of time spent in processing. Figure~\ref{fig:noexp} shows the percentage of time taken by Naive Bilateral filtering algorithm\cite{Tomasi1998} and an improved algorithm employing tiled-memory-access pattern on different platforms for a 512x512 pixels small image and a 3000x3000 pixels large image after the $exp$ function was commented out from the computation. The only purpose of this experiment is to justify the fact that traditional memory-access-pattern related optimizations can not improve the performance of Bilateral filtering kernel.

@@

Since the $exp$ instruction does not involve transfer of data from main memory, it follows from this experiment that memory reuse and memory access pattern related optimizations are not of much help to speed up Bilateral filtering kernel. What Amdahl's law claims, for inherently sequential portion of a parallel algorithm, can be applied here. \\ \textit{The total speedup achieved by optimizations that deal with memory management is limited by the amount of time spent in memory transfers}. Therefore, in this case, traditional optimizations such as tiling/blocking, cache efficient initializations, prefetching, and other NUMA optimizations can only result in a speed up of at most 2\% and hence we have concentrated on optimizations that deal with the complexities of computational aspects of Bilateral filtering kernel.

@@

Our CPU based implementation of Bilateral filtering algorithm explores the use of special purpose registers using both assembly level programming and SSE intrinsics. The advantage of using assembly language is that a programmer can have explicit control of underlying hardware. However, programmer productivity is affected and an in-depth knowledge of underlying hardware is essential.

@@

\subsection{Reduction Methods}

\label{sub:reduction} Sum by reduction is a technique to compute sum of multiple elements usually stored in contiguous memory locations. There are assembly instructions that can add horizontally packed, single precision, floating point numbers. We used $haddps$ assembly instruction that takes two (source and destination) 128 bit operands, and adds the single precision floating point numbers stored in first and second $dword$ of destination and source operand and stores it in first and third $dword$ of the destination operand respectively. Similarly, the third and fourth $dword$ of destination and source operand are added and stored in second and fourth $dword$ of the destination operand respectively. Both source and destination values must be stored in MMX registers. The SSE intrinsic to perform horizontal add is $\_mm\_hadd\_ps$. It takes two 128 bit operands, each of which is treated as four 32-bit floating point elements. The result of this operation on operand X(X0, X1, X2, X3) and Y(Y0, Y1, Y2, Y3) is (Y3 + Y2, Y1 + Y0, X3 + X2, X1 + X0).

@@

\begin{table*}[t]
\begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{3.5cm}|l|p{2cm}|l|p{0.9cm}|}
\hline 
\textbf{Model}  & \textbf{Clock frequency}  & \textbf{cores per chip}  & \textbf{chips per node}  & \textbf{cores per node}  & \textbf{SIMD support}  & \textbf{Cache/Memory}  & Compiler\\
\hline 
Intel Xeon E5410  & 2.33 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 12 MB  & icpc 11.1\\
\hline 
Intel Xeon X7560  & 2.27 $GH_z$  & 32  & 1  & 32  & SSE, SSE2  & 24 MB & icpc 11.1\\
\hline 
Intel Core i7-870  & 2.93 $GH_z$  & 4  & 1  & 4  & SSE, SSE2, SSE3, SSE4.2  & 8MB & icpc 12.0.2\\
\hline 
AMD Opteron 2376  & 2.3 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 6MB  & icpc 11.1\\
\hline 
AMD Opteron 8350  & 2.0 $GH_z$  & 4  & 4  & 16  & SSE, SSE2  & 2MB  & icpc 11.1 \\
\hline 
AMD Phenom II X6 1045T  & 2.7 $GH_z$  & 3  & 1  & 6  & SSE, SSE2, SSE4a  & 6MB  & icpc 12.0.2 \\
\hline 
NVidia GeForce GTX 280 & 1296 MHz & 240 & 1 & 240 & - & 1 GB GDDR3 & CUDA 2.0\\
\hline
\end{tabular}\caption{Architectural details of multicore chips employed for experiments}
\label{tab:archs} 
\end{table*}

In order to evaluate our implementation we have tried it on a range of multicore and manycore chips. Table \ref{tab:archs} summarizes architectural details of systems we used during our experiments.

>>>>

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images. As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e. one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:
\begin{equation}
\label{eq:depen}
filtered\: pixel\: x=\frac{\substack{N\\
\sum\\
i=1
}
f(\xi_{i})\cdot c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}{\substack{N\\
\sum\\
i=1
}
c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}
\end{equation}

In addition to restating the bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the bilateral filter. Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated. Consequently, the bilateral filter has two data dependencies. As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.

Another observation taken into account is that the values of Equation~\ref{spatial}, for the Neighborhood, can be computed in advance as the values depend on a simple Gaussian probability distribution function. However, the values of Equation~\ref{photometric} and consequently of Equation~\ref{combined} cannot be known a priori as they depend on the similarity of signals between the current Center Pixel and the Neighborhood. Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window. By reason of the foregoing, the bulk of the time spent in processing an image using the bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.

\subsection{A Naive Approach to bilateral filtering kernel}

\begin{figure}
\centering\includegraphics[scale=0.35]{images/the1}
\includegraphics[clip,scale=0.35]{images/the202}\includegraphics[scale=0.35]{images/the2}\includegraphics[scale=0.35]{images/the3}
\caption{Example of a naive bilateral filter iterations}
\end{figure}

The naive stencil algorithm is the standard model that is predominantly used in bilateral filtering kernel. Previous implementations\cite{Paris2009,Pham2005} of bilateral filtering kernel improves the hardware dependent aspects and hence expend resources on achieving performance gains that can not match the effort taken. Here, we address a root problem of the standard naive algorithm by proposing a new algorithm, and then develop hardware dependent improvements for further speedup. Algorithm~\ref{alg:Example-Algorithm-float} is the pseudo-code for the standard naive bilateral filtering kernel.

\begin{algorithm}[!h]
\caption{Naive bilateral Filter Algorithm}
\label{alg:Example-Algorithm-float}
\begin{algorithmic}
\FORALL{element  $s_{i,j}$ in source image s}
\STATE{$X_{1} \leftarrow 0$}
\STATE{$X_{2} \leftarrow 0$}
\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\ENDFOR
\ENDFOR
\STATE{output[$i,j] \leftarrow X_{1} / X_{2}$}
\ENDFOR
\end{algorithmic}
\end{algorithm}

\subsection{Pair-symmetric bilateral filtering kernel}
The pair-symmetric algorithm, shown in Algorithm~\ref{alg:Example-Algorithm-float-1}, reduces the amount of Center-Neighbor computations by half. Thus, it can theoretically cut the processing time in half at the cost of increased memory requirement, which increases with the amount of dependencies. Its implementation is non-trivial due to potential race-conditions.

Our pair-symmetric bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel, and that the codes are symmetric with respect to Center-Neighbor paired computations. For example, in the bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$. It then follows that the naive bilateral filtering algorithm dictated in Algorithm~\ref{alg:Example-Algorithm-float} above, although simple, performs twice as many calculations than necessary.

The Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive bilateral filtering algorithm. Instead of reading from all Neighbors, our pair-symmetric algorithm reads from only half of the Neighbors. The result of each Center-Neighbor paired computation is then added to the $X_n$ register, which eventually obtains the result of data dependency $n$. This is trivial as the naive bilateral filtering algorithm already does this. What our pair-symmetric algorithm does unique is that the result of each Center-Neighbor paired computation is also added to and stored in the Neighbor's position in a temporary array ("Map'') that is used for storing the results of a data dependency. So, there are $n$ Maps for $n$ number of dependencies.

After the data dependency conditions are met, the function that returns the output data from the dependencies iterates over the Maps. As applied to the bilateral filter, the data dependency lies in calculating each pixel's numerator and denominator that is used to derive the pixel's output result. After the numerators and denominators are derived, a function iterates and divides the corresponding elements of the two arrays to return the output.

@@

In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.

\begin{figure}
\subfigure[Each warp is assigned a row in filter window]{\includegraphics[width = 0.48 \textwidth]{images/project24.pdf}
}\label{fig:pspro}
\subfigure[2 x 2 thread-blocks each with 4 x 3 threads]{\includegraphics[width = 0.48  \textwidth]{images/tiles19.pdf}
}\label{fig:pspro2}
\subfigure[An illustrated overview of the bilateral filtering process]{\includegraphics[width = 0.48  \textwidth]{images/smallfaces.pdf}}
\label{fig:pspro3}
\caption{Memory access patterns for CUDA that exploit implicit intra-warp synchronization}
\label{fig:intrawarp}
\end{figure}

\subsection{Shared Memory and Tile Processing}

CUDA features three types of memory with read/write capability suitable for general-purpose computing on GPU: \begin{inparaenum}[(i)] \item global memory, which is accessible by all threads in all thread blocks, \item shared memory, which is provided separately for each block of threads to share among themselves, and \item registers, which are separate and distinct for each thread and of which each is only accessible by the thread to which it belongs\end{inparaenum}. Of these three types of memory, the slowest is global memory, which unfortunately is where the host must first send the input data before the GPU can begin its work. So, in order to see any meaningful performance in CUDA, data stored in global memory must be transferred to faster memory, i.e. shared memory, which is hundreds of times faster. However, the trade-off is that the shared memory size is much smaller than the global memory size, and this fact compels a tile-based approach to the implementation of bilateral filtering algorithm.

Figure~\ref{fig:Tiling-for-Pair-Symmetric} illustrates how tiles are processed in parallel without explicit mutual exclusion. 
\begin{wrapfigure}{r}{0.25\textwidth}
\centering
\includegraphics[width=0.24\textwidth]{images/stackedtiles}
\caption{Tiling for Pair-Symmetric bilateral filtering with each tile having a unique color. Overlapping tiles have a resultant color.}
\vspace{0mm}
\label{fig:Tiling-for-Pair-Symmetric}
\end{wrapfigure}
In tile based algorithm, equal sized partitions of the input data are processed, and then any remaining data on and between the edge and the last partition that is closest to (but not falling over) the edge boundary of the 2-dimensional data representation is processed. Each thread only updates (writes to) only half of the Neighbors. Due to this access pattern, greater number of threads can work in parallel. 
With respect to the naive bilateral filtering algorithm, tile-based processing is trivial and needs no further explanation because simply partitioning the space and processing it will suffice. However, for the pair-symmetric algorithm that we propose, tile-based bilateral filtering algorithm requires adjustments in the order in which tiles are processed in order to guarantee the exclusion of race conditions.

\begin{wrapfigure}{r}{0.25\textwidth}
\includegraphics[width=0.24\textwidth]{images/lenatiles}
\caption{Input image after one iteration of pair-symmetric algorithm. Tiles do not have perfect boundaries due to overlaps.}
\label{fig:An-iteration-of}
\end{wrapfigure}
If processes were assigned to adjacent tiles, race conditions would occur because the edge cells of the tiles would overlap each other due to intersecting half-width periphery regions, as illustrated in Figure~\ref{fig:Tiling-for-Pair-Symmetric}. To get around this problem, simultaneously processed tiles can be spaced apart in strides such that no two tiles being processed at the same time intersect with one another. An example of such a technique is illustrated in Figure~\ref{fig:An-iteration-of} - the tiles are flipped sideways in this rendering.

@@

\subsection{Memory to computation ratio of bilateral filtering kernel}
\begin{figure}[h!]
\begin{centering}
\includegraphics[width=0.48\textwidth]{images/noexplarge}
\end{centering}
\caption{Percentage of time taken by sequential kernel when run without $exp$ instruction, $filter$ $radius$=10.}
\label{fig:noexp} 
\end{figure}

Unlike traditional regular embarrassingly parallel stencil kernel codes where memory management plays a pivotal role in realizing performance, bilateral filtering kernel is extremely computationally intensive with only a meager amount of time spent in memory accesses. Bilateral filtering kernel requires computation of $exponential$ function ($exp$ function), as shown in Algorithm~\ref{alg:Example-Algorithm-float}, which constitutes a major component of time spent in processing. Figure~\ref{fig:noexp} shows the percentage of time taken by Naive bilateral filtering algorithm and a modified algorithm, employing tiled-memory-access pattern, on different multicore platforms for a 512x512 pixels and another 3000x3000 pixels image, after the $exp$ function was commented out from the algorithms. The only purpose of this experiment is to justify the fact that traditional memory-access-pattern related optimizations can not improve the performance of bilateral filtering kernel.

@@

Since the $exp$ instruction does not involve transfer of data from main memory, it follows from this experiment that memory reuse and memory access pattern related optimizations are not of much help to speed up bilateral filtering kernel. What Amdahl's law claims, for inherently sequential portion of a parallel algorithm, can be applied here. \\ \textit{The total speedup achieved by optimizations that deal with memory management is limited by the amount of time spent in memory transfers}. Therefore, in this case, traditional optimizations such as tiling/blocking, cache efficient initializations, prefetching, and other NUMA optimizations can only result in a speed up of at most 2\% and hence we have concentrated on optimizations that deal with the complexities of computational aspects of bilateral filtering kernel.

\begin{table*}[t!]
\begin{tabular}{|l|p{1.2cm}|p{0.8cm}|p{0.8cm}|p{0.8cm}|p{3.5cm}|l|p{2cm}|l|p{0.9cm}|}
\hline 
\textbf{Model}  & \textbf{Clock frequency}  & \textbf{cores per chip}  & \textbf{chips per node}  & \textbf{cores per node}  & \textbf{SIMD support}  & \textbf{Cache/Memory}  & Compiler\\
\hline 
Intel Xeon E5410  & 2.33 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 12 MB  & icpc 11.1\\
\hline 
Intel Xeon X7560  & 2.27 $GH_z$  & 32  & 1  & 32  & SSE, SSE2  & 24 MB & icpc 11.1\\
\hline 
Intel Core i7-870  & 2.93 $GH_z$  & 4  & 1  & 4  & SSE, SSE2, SSE3, SSE4.2  & 8MB & icpc 12.0.2\\
\hline 
AMD Opteron 2376  & 2.3 $GH_z$  & 4  & 2  & 8  & SSE, SSE2  & 6MB  & icpc 11.1\\
\hline 
AMD Opteron 8350  & 2.0 $GH_z$  & 4  & 4  & 16  & SSE, SSE2  & 2MB  & icpc 11.1 \\
\hline 
AMD Phenom II X6 1045T  & 2.7 $GH_z$  & 3  & 1  & 6  & SSE, SSE2, SSE4a  & 6MB  & icpc 12.0.2 \\
\hline 
NVidia GeForce GTX 280 & 1296 MHz & 240 & 1 & 240 & - & 1 GB GDDR3 & CUDA 2.0\\
\hline
\end{tabular}
\caption{Architectural details of multicore chips employed for experiments}
\label{tab:archs} 
\end{table*}

@@

Our CPU based implementation of bilateral filtering algorithm explores the use of special purpose registers using both assembly level programming and SSE intrinsics. The advantage of using assembly language is that a programmer can have explicit control of underlying hardware. However, programmer productivity is affected and an in-depth knowledge of underlying hardware is essential.

@@

\subsection{Reduction Methods}
\label{sub:reduction} Sum by reduction is a technique to compute sum of multiple elements usually stored in contiguous memory locations. There are assembly instructions that can add horizontally packed, single precision, floating point numbers. We used $haddps$ assembly instruction that takes two (source and destination) 128 bit operands, and adds the single precision floating point numbers stored in first and second $dword$ of destination and source operand and stores it in first and third $dword$ of the destination operand respectively. Similarly, the third and fourth $dword$ of destination and source operand are added and stored in second and fourth $dword$ of the destination operand respectively. Both source and destination values must be stored in MMX registers. The SSE intrinsic to perform horizontal add is $\_mm\_hadd\_ps$. It takes two 128 bit operands, each of which is treated as four 32-bit floating point elements. The result of this operation on operand X(X0, X1, X2, X3) and Y(Y0, Y1, Y2, Y3) is (Y3 + Y2, Y1 + Y0, X3 + X2, X1 + X0).

@@

In order to evaluate our implementation we have tried it on a range of multicore and manycore chips. Table \ref{tab:archs} summarizes architectural details of systems we used during our experiments. Our implementation for multicores is written in $C++$ and uses OpenMP for parallelization; the GPU implementation uses CUDA. To enrich supported image formats, we use $CImg$ library.
